---
title: "PSPG 245A: PK concepts with PKPDsim - Workshop 1 Tasks"
author: "Savic Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: TRUE
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

<br> <br> <b> Exercises </b><br>

1) IV 1-CMT PK MODEL

a) Build an IV 1-cmt PK model

b) Simulate three doses: 10 mg, 100 mg, 1000 mg

c) How does a 100-fold increase in CL affect the curve? What about
100-fold decrease in V?

2) IV 2-CMT PK MODEL

a) Build an IV 2-cmt PK model

b) Simulate multiple doses

c) BONUS: How does a 100-fold increase in V2 affect the curve?

**PKPDsim documentation:**
<https://github.com/ronkeizer/pkpdsim-docs/tree/master>

# Installing and Loading packages

```{r warning=F, message=F}
# To install, remove the # before command and run the line (CTRL + RETURN)
# Comment out (with #) if you've already installed them

# To install PKPDsim
# library(devtools)
# install_github("InsightRX/PKPDsim")

# To install other packages that we'll use today
# install.packages("dplyr")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("NonCompart")

# Load libraries
library(PKPDsim)
library(dplyr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(NonCompart)
```

# Workshop #1

## Model components

Running a simulation requires THREE components: <br> 1. Model <br> 2.
Parameters <br> 3. Regimen <br>

## EXERCISE 1. IV 1-CMT PK MODEL

### a) Build an IV 1-cmt PK model

#### 1. Model

```{r}

# The PKPDsim library has pre-packaged models. 
# They can be viewed by running new_ode_model() with an empty bracket. 
# You can deploy them by saving to a variable. In this example, we'll save to a variable called "my_model":

# i.e. my_model <- new_ode_model("pk_1cmt_oral")

# Here we will hard code the model for an IV  1-cmt model. 

my_model <- new_ode_model(code = "
                          dAdt[1] =  (-CL/V) * A[1]
                          ",
                          obs = list(cmt = 1, scale = "V"),
                          declare_variables = c("CP"))

# run "my_model" to see the model summary and required parameters

my_model

```

#### 2. Parameters

```{r}
# Our drug of interest has a volume of 250 L and clearance of 10 L/h.

# Save your PK parameters as variables here:

V = 250 # Central volume (L)
CL =  10 # Clearance (L/h)


# We'll store these in a list that PKPDsim can read
my_parameters <- list("V"  = V,
                      "CL" = CL
                      )

```

#### 3. Regimen

```{r}
# Let's simulate one single dose of 100 mg at time 0
# To prescribe a regimen, we need to use the PKPDsim function, new_regimen()
# Look up ?new_regimen to see arguments this function takes

?new_regimen

my_regimen <- new_regimen(
  amt = 100,
  times = c(0)  
)

```

#### 4. Simulate!

```{r}
# Put it all together 
# Use sim() to create a dataset of simulation

# fill in the three arguments with our three components from above
my_simulated_df_d100 <- sim(ode = my_model,
                       regimen = my_regimen,
                       parameters = my_parameters,
                       only_obs = T,
                       output_include = list(variables=T)
                       )
# Open your new dataset and understand the set up. 
# What is each column? What are the units of each column?
my_simulated_df_d100 %>% data.frame %>% colnames()
# What is the concentration at time 0? What is another way we could have calculated C(0)?
# concentration -> y
# so C(0)=0.4
my_simulated_df_d100 %>%
  filter(t==0) %>%
  t
```

#### 5. Plot

```{r}
# use ggplot to plot the data 

my_simulated_df_d100 %>% 
  ggplot(aes(x=t, y=y)) +
  geom_point()
```

### b) Simulate different doses: 10 mg , 100 mg, 1000 mg

Plot all simulations on one plot.

#### New Regimen

```{r}
# Add two new regimens of one single dose of either 10 mg or 1000 mg at time 0

my_regimen_d10 = new_regimen(
  amt = 10, 
  time = c(0)
)

my_regimen_d1000 = new_regimen(
  amt = 1000, 
  time = c(0)
)

```

#### Simulate

Create a new simulation for each new regimen HINT: use mutate to add a
column to identify the simulationa and rbind the simulations together

```{r}
# Put it all together
# Use sim() to create a dataset of simulation
my_simulated_df_d10 = sim(
  ode = my_model,
  parameters = my_parameters,
  regimen = my_regimen_d10,
  only_obs = T,
  output_include = list(variables=T)
  
)

my_simulated_df_d1000 = sim(
  ode = my_model,
  parameters = my_parameters,
  regimen = my_regimen_d1000,
  only_obs = T,
  output_include = list(variables=T)
)

my_simulated_df = rbind(
  my_simulated_df_d100 %>% mutate(dose='100mg'),
  my_simulated_df_d10 %>% mutate(dose='10mg'),
  my_simulated_df_d1000 %>% mutate(dose='1000mg')
)
```

#### Plot

```{r}
# plot the data with ggplot
# what does the plot look like on the semi-logarithmic scale?
# what are the units of t and y?
# can you add labels to the axes?

my_simulated_df %>%
  ggplot(aes(x=t, y=y, colour=dose)) +
  geom_point() +
  scale_y_log10() + 
  labs(x = "Time (h)", y="Conc. (mg/L)")
```

### c) How does a 10-fold increase in CL affect the curve? What about 10-fold decrease in V?

#### Update parameters

```{r}
# Save your PK parameters as variables here:
# Don't forget to store them in a list for PKPDsim.

my_parameters_10_fold_CL <- list(
  "V"  = V,
  "CL" = CL * 10
)


my_parameters_10_fold_V <- list(
  "V"  = V / 10,
  "CL" = CL
)
```

#### Simulate

```{r}
# Put it all together
# Use sim() to create a dataset of simulation

my_simulated_df_change_CL_or_V = rbind(
  sim(
    ode = my_model,
    parameters = my_parameters_10_fold_CL,
    regimen = my_regimen,
    only_obs = T,
    output_include = list(variables=T)
    
  ) %>% mutate(label='CL * 10'),
  
  sim(
    ode = my_model,
    parameters = my_parameters_10_fold_V,
    regimen = my_regimen,
    only_obs = T,
    output_include = list(variables=T)
  ) %>% mutate(label='V / 10')
)
```

#### Plot

Plot all simulations on one plot.

```{r}
# plot the data with ggplot
my_simulated_df_change_CL_or_V %>%
  ggplot(aes(x=t, y=y, colour=label)) +
  geom_point() +
  #scale_y_log10() +
  labs(
    x = "Time (h)", 
    y = "Conc. (mg/L)"
  )
```

## EXERCISE 2. IV 2-CMT PK MODEL

### a) Build an IV 2-cmt PK model

#### 1. Model

```{r}

# Write ODEs for an IV 2-cmt model

my_model_2cmt <- new_ode_model(code = "
                               dAdt[1] = -CL/V*A[1] - Q/V*A[1] + Q/V2*A[2]
                               dAdt[2] = Q/V*A[1] - Q/V2*A[2]
                               CP = A[1]/V
                               ",
                               obs = list(cmt = 1, scale = "V"),
                               declare_variables = c("CP")
                               )

# run "my_model" to see the model summary and required parameters

my_model_2cmt

```

#### 2. Parameters

```{r}

# For our two compartment, we have two more parameters: 1) the intercompartmental clearance (Q) and peripheral volume (V2).
# Let's set the intercompartmental clearance to 25 L/h.
# For volume, let's make the total volume 250 L, where the central volume is 25 L. (V_TOTAL = V + V2)

# Save your PK parameters as variables here:

my_parameters_2cmt = c(
  CL = 10,
  V  = 25,
  V2 = 250 - 25,
  Q  = 25
)
```

#### 3. Regimen

```{r}
# One single dose of 100 mg at time 0
# Look up ?new_regimen to see arguments this function takes

my_regimen_2cmt = new_regimen(
  amt = 100,
  times = 0,
)
```

#### 4. Simulate

```{r}
# Put it all together <br>
# Use sim() to create a dataset of simulation

my_simulated_df_2cmt = sim(
  ode = my_model_2cmt,
  parameters = my_parameters_2cmt,
  regimen = my_regimen_2cmt,
  only_obs = T,
  output_include = list(variables=T)
  
)
```

#### 5. Plot

```{r}
# use ggplot to plot the data 
my_simulated_df_2cmt %>% 
  ggplot(aes(x=t, y=y)) +
  scale_y_log10() +
  geom_point()
```

### b) Simulate multiple doses

#### Regimen

```{r}
# Instead of a single dose, now let's look at multiple doses across 7 days

# HINT: for multiple dosing, you will need a vector of times.
# Don't forget what our units for time are!

# A vector can be created using c(). 
## ex. c(0,24,48,72)

# Another way is to use seq(). 
## ex. seq(start_time, end_time, increment)
## ex. daily dosing for 28 days: seq(0,672,24)

my_regimen_multiple <- new_regimen(
  amt = 100,
  times=seq(0,24*7,24)
)
```

#### Simulate

```{r}
# Put it all together
# Use sim() to create a dataset of simulation


my_simulated_df_multiple = sim(
  ode = my_model_2cmt,
  parameters = my_parameters_2cmt,
  regimen = my_regimen_multiple,
  only_obs = T,
  output_include = list(variables=T)
  
)
```

#### Plot

```{r}
# use ggplot to plot the data 
my_simulated_df_multiple %>% 
  ggplot(aes(x=t,y=y)) + 
  geom_point() + 
  geom_line() 

```

### c) BONUS: How does a 100-fold increase in V2 affect the curve?

#### Update parameters

```{r}
# Save your PK parameters as variables here:

my_parameters_2cmt_100_fold_V2 = list(
  CL = 10,
  V  = 25,
  V2 = (250 - 25) * 100,
  Q  = 25
)
```

####  Regimen

```{r}
# keep same multidose regimen: 100 mg once daily for 7 days

my_regimen_multiple
```

#### Simulate

```{r}
# Put it all together 
# Use sim() to create a dataset of simulation

my_simulated_df_multiple_high_V2 = sim(
  ode = my_model_2cmt,
  parameters = my_parameters_2cmt_100_fold_V2,
  regimen = my_regimen_multiple,
  only_obs = T,
  output_include = list(variables=T)
  
)
```

#### Plot

```{r}
# plot the data with ggplot
ggplot() + 
  geom_point(data=my_simulated_df_multiple, aes(x=t, y=y), color ="red") + 
  geom_line(data=my_simulated_df_multiple, aes(x=t, y=y), color ="red") + 

  geom_point(data=my_simulated_df_multiple_high_V2, aes(x=t, y=y), color ="blue") + 
  geom_line(data=my_simulated_df_multiple_high_V2, aes(x=t, y=y), color ="blue") + 

  scale_y_log10()

```

What do you notice with these two curves?
