---
title: "PSPG 245A: PK concepts with PKPDsim - Workshop 3 Tasks"
author: "Savic Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float: TRUE
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

<br> <br> <b> Exercises </b><br>

<b> 5) VARIABILITY part 1 </b> <br> a) Simulate 200 patients including
IIV in the model <br> b) Re-visit optimal dosing if needed

<b> 6) VARIABILITY part 2 </b> <br> a) Simulate 200 patients for each WT
group including IIV <br> b) Re-visit optimal dosing if needed

<b> PKPDsim documentation: </b> <br>
<https://github.com/ronkeizer/pkpdsim-docs/tree/master> <br>

# Installing and Loading packages

```{r warning=F, message=F}
# To install, remove the # before command and run the line (CTRL + RETURN)
# Comment out (with #) if you've already installed them

# To install PKPDsim
# library(devtools)
# install_github("InsightRX/PKPDsim")

# To install other packages that we'll use today
# install.packages("dplyr")
# install.packages("tidyr")
# install.packages("ggplot2")
# install.packages("NonCompart")

# Load libraries
library(PKPDsim)
library(dplyr)
library(tidyr)
library(ggplot2)
library(NonCompart)

```

# Workshop #3

## EXERCISE 5: Simulate 200 patients including IIV in the model and re-visit optimal dosing if needed <br>

NEW: For the compound, savic-245a, the efficacy and toxicity thresholds
were associated with Cmin,ss, or the minimum concentration at the end of
a dosing interval at steady state. <br>

<br> Thresholds are: <br> Cmin,ss efficacy= 25 mg/L <br> Cmin,ss
toxicity= 60 mg/L

<br> Variability (coefficient of variation (CV)) associated with V & CL
<br> V- CV 10% <br> CL- CV 15%

### a.) Simulate the oral 1-cmt model from the CHALLENGE QUESTION with Inter Individual Variability

#### 1. Model

```{r}
# write ODEs for an oral dose, 1 compartment model

my_model <- new_ode_model(code = "
                          dAdt[1] = -KA*A[1]
                          dAdt[2] = KA*A[1] - CL/V*A[2]
                          CP = A[2]/V
                          ",
                          obs = list(cmt = 2, scale = "V"),
                          declare_variables = c("CP"))

# run "my_model" to see the model summary and required parameters
my_model

```

#### 2. Parameters

```{r}
# Input your PK parameters here:

V = 30 # Central volume (L)
CL = 0.8 # Clearance (L/h)
KA = 1 # Rate of absorption (1/h)

# We'll store these in a list that PKPDsim can read
my_parameters <- list("V"  = V,
                      "CL" = CL,
                      "KA" = KA
                      )
  
```

#### 3. Regimen

```{r}
# 900 mg QD for 14 days
my_regimen <- new_regimen(
  amt = 900, # 900 mg
  times = seq(0,14*24,24)
)
```

#### 4. Simulate!

Put it all together <br> Use sim() to create a dataset of simulation

```{r}
# to simulate between-subject variability, we need to add another argument, omega.
# we are given coefficient of variation (CV), what's the relationship between variance (omega) and CV?
# PKPDsim (and other PK packages) require omegas in a lower diagonal matrix to account for covariance. 
# If you don't have covariance, the off-diagonal values are 0.
# to make a vector, take the lower diagonal matrix and read top-bottom, left-right.
#
# ex.
# 1
# 2 3
# 4 5 6
#
# = c(1,2,3,4,5,6)

# fill in ode, regimen, parameters
# to simulate multiple patients, we need to add an argument. Can you find it in the sim() documentation?


my_simulated_df = sim(
  ode = my_model,
  regimen = my_regimen,
  parameters = my_parameters,
  omega = cv_to_omega(
    par_cv = list(V = .10, CL = .15, KA = 0), # V- CV 10%, CL- CV 15%
    my_parameters
  ),
  t_obs = seq(0,24*14,1), # how freq. you want to see your observation, e.g. c(0,24,1) ... every hour 
  
  n_ind = 200,
  seed = 123,
  
  only_obs = T,
  output_include = list(variables = T, parameters = T)
)

my_simulated_df
```

#### 5. Plot

```{r}
# use ggplot to plot the data
# let's save our efficacy and threshold values as variables

eff_threshold = 25 # mg/L
tox_threshold = 60 # mg/L

my_simulated_df %>% 
  ggplot() +
  geom_point(aes(x = t, y = y)) + 
  geom_hline(yintercept = tox_threshold, colour = 'red', size = 1.2, alpha = .5) +
  geom_hline(yintercept = eff_threshold, colour = 'blue', size = 1.2, alpha = .5)
```

### b.) Do we have at least 90% of patients within the defined therapeutic window? <br>

Cmin,ss efficacy= 25 mg/L <br> Cmin,ss toxicity= 60 mg/L <br>

```{r}
# Select cmin at ss for each id

# To do this, create a new dataset based off your simulation dataset that has one row per id and the concentration at the end of the dosing interval.

# Now make a new column that categorizes the patients into above or below threshold with a "Yes" or "No"
# You can do this using mutate and an ifelse() statement.

my_simulated_df %>% 
  filter(t == max(t)) %>%
  mutate(ifelse())

# make a table of the Therapeutic window result
# take to Google and learn how you can make the table a proportion of the total 

```

## EXERCISE 6: Simulate 200 patients for each WT group including IIV, and re-visit optimal dosing if needed

An increase in CL has been associated with weight (WT) for savic-245a.
Different typical clearances have been determined for low and high body
WT.

Low body weight CL = 0.4 L/h High body weight CL = 1.04 L/h

<br> Because we've used weight to describe variability in CL, the CV
associated with CL has decreased. <br> V- CV 10% <br> CL- CV 8%

### a.) Simulate the PK profile for 200 patients with low WT and for 200 patients with high WT

#### 1. Model

```{r}
# write ODEs for an oral dose, 1 compartment model

my_model <- new_ode_model(code = "
                          dAdt[1] = -KA*A[1]
                          dAdt[2] = KA*A[1] - CL/V*A[2]
                          CP = A[2]/V
                          ",
                          obs = list(cmt = 2, scale = "V"),
                          declare_variables = c("CP"))

# run "my_model" to see the model summary and required parameters
my_model

```

#### 2. Parameters

```{r}
# Input your PK parameters here:


# We'll store these in a list that PKPDsim can read

 
```

#### 3. Regimen

```{r}
# 900 mg QD for 14 days

```

#### 4. Simulate!

Put it all together <br> Use sim() to create a dataset of simulation

```{r}
# fill in the three arguments with our three components from above
my_simulated_df_lowWT <- 

my_simulated_df_hiWT <- 


```

#### 5. Plot

```{r}
# use ggplot to plot the data, one per each weight group


```

### b.) Do we have at least 90% of patients within the defined therapeutic window? <br>

Cmin,ss efficacy= 25 mg/L <br> Cmin,ss toxicity= 60 mg/L <br>

```{r}
# Select cmin at ss for each id
# What is the proportion of patients within the defined therapeutic window for each weight group?

```

### c.) Change the dosing schedules to have at least 90% of the patients within the defined therapeutic window <br>

Available formulations: <br> -300 & 150 mg pills <br> (can be combined)
<br>

#### 1. Regimen

```{r}

```

#### 2. Simulate!

```{r}

```

#### 3. Plot

```{r}

```

#### Do we have now at least 90% of patients withing the defined therapeutic window? <br>

Cmin,ss efficacy= 25 mg/L <br> Cmin,ss toxicity= 60 mg/L <br>

```{r}

```
